Think deep for reef resilience

### 2025-09-03 paper/paper_draft.qmd
Paper updates from comments
- move to 'intervention' instead of 'restoration'
- Expand more on connectivity interactions with depth and expand on the findings relating to the most suitable targeting strategies for further investigation (need to add in input maps)
- Add section about the need for connectivity model comparisons as well as validation of models and increased yearly sample sizes to improve confidence in connectivity modelling

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the performance of deep reefs, contrasting current restoration
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as deployment of coral seeding and fragment devices, target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit justifications for this depth range, restoration depth is likely constrained by snorkelling and diving deployment and monitoring logistical accessibility, visual identification of sites from the surface, in combination with tourism return value. Our results suggest that of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region scale), with many unable to maintain positive carbonate budgets under future conditions. Partial dependence analysis revealed a range of depths from 10-15m where the probability of a reef being clustered into the High cover cluster increased substantially. Additionally, two-way partial dependence analysis revealed that high incoming connectivity is important for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth range, and targetting well connected reefs, when assessing possible restoration sites may reduce the exposure of planted and seeded corals to high thermal stress, while still maintaining effective logistical access to sites. Managers are also likely to consider the positive potential for flow on effects due to outgoing connectivity when selecting intervention sites [@hock2017]. While reefs with high outgoing connectivity levels were found to have greater probability of being assigned to low cover clusters when assessed individually (@fig-ts-cluster-pdp-twoway-incoming), within the context of reef depth, reefs with high depths and high outgoing connectivity levels had the greatest probability of being clustered into high cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results suggests that interventions targetting deep reefs will also benefit from high incoming and outgoing connectivity levels, allowing for high potential for site resilience and outgoing flow of benefits and recovery. Transient, and difficult to measure, properties such as reef connectivity are likely to increase uncertainty in intervention planning to a greater degree than environmental characteristics such as reef depth. Overall, reef intervention planning is a complex and challenging decision-making process [@anthony2020], and our results suggest that intervention site depth may be an important factor to additionally consider when assessing the long-term resilience of sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is the degree of larval connectivity between reefs and the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008; @torres2024]. In contrast, our results suggest that while connectivity metrics show some levels of cluster predictor importance, there are few differences between resilient and non-resilient reefs in metrics such as source-to-sink connectivity ratio and total connectivity strength. While there is interaction between incoming connectivity and depth in reef cluster prediction, according to available input data, there are few reefs in the GBR that occur at depths \>10m and have Log10 weighted incoming connectivity levels below 0 (@sfig-access-esm1-5-mean-carb-budget), resulting in low importance of connectivity as an individual feature for clustering across the GBR. Furthermore, our results suggest no relationship between connectivity strength and the ability of reefs to maintain positive carbonate budgets under future conditions. There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling is able to identify reefs that are important for recovery processes, with findings dependent on the larval connectivity model used. These investigations were performed using prior generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods, without considering the influence of reef depth on resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a limited temporal sample of particle flows, our results suggest that larval connectivity alone may not be important or reef resilience and recovery, but should be considered in combination with more influential factors such as depth.

Our findings on the importance of connectivity may differ due to the highly uncertain nature of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching events decreasing the breeding capacity of strongly connected reefs. Analysing daily connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for capturing the seasonal - ecological time scale connectivity between locations due to high daily variability in connectivity data. This is supported by the connectivity data used in our investigation, which have large relative standard deviations for connectivity between reefs over the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit connectivity [@vogt-vincent2023], future investigations may find increased importance of both depth and reef connectivity for resilience, in line with other investigations [@hock2017; @torres2024]. Further improvements and research in connectivity modelling are necessary when incorporating connectivity as a supplementary factor in reef resilience and intervention-planning analyses.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced available light levels. [@baker1975; @huston1985; @ruiz-diaz2022]. If included in this investigation, it is possible that reef recovery ability would be reduced at greater depths, reducing the positive effect of depth refugia for temporal resilience. Additionally, reduced growth rates at increased depths may reduce the carbonate budget maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress projections are likely to influence the results obtained in this investigation. While the resolution of connectivity data used in this investigation improves upon data used in comparable studies [@bozec2025], the time-mean implementation used may not effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity implementation methods that better capture the ecological-timescale trends in reef connections rather than the high daily variability, would improve confidence in the connectivity findings of this investigation [@vogt-vincent2023]. The DHW data used in this investigation have not been bias corrected for the Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022] developed using statistical downscaling. Additionally, Lachs et al. [@lachs2021] suggest the utility of DHW levels for prediction of coral bleaching can be improved by reducing the threshold for HotSpot Anomaly values from MMM +1 to MMM and reducing the accumulation window from 12 weeks to 8 weeks. We investigate a range of future climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox to account for many possible futures, and our results suggest the factors that influence reef resilience are consistent across these scenarios.

@@

Improvements in remote sensing approaches for broad scale bathymetry measurement and understanding of how depth impacts experienced thermal stress on reefs will further enhance investigations of depth as a critical factor for reef resilience. Additional analyses could build on the findings of this study by assessing the impacts of prioritising reef depth in reef restoration decision making. Future analyses could incorporate other environmental and ecological stressors such as wave activity, ocean acidification, and Crown-of-Thorn Starfish [@lin2023; @mcclanahan2005; @grimaldi2023; @grimaldi2022; @liu2024; @eyre2018].

@@

The results presented suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of thermal refugia for shallow-water reefs on the GBR. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts [@anthony2020; @hock2017], current connectivity modelling methods, in combination with our modelling framework, suggest that connectivity may not be as solely-important for reef resilience under projected conditions as environmental factors such as depth and thermal stress levels. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef intervention decision making, as well as further research and development in connectivity modelling methods for whole-ecosystem scales.

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Current reef restoration and intervention investigation efforts on the GBR, such as deployment of coral seeding and fragment devices, target reefs with suitable sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit justifications for this depth range, intervention depth is likely constrained by snorkelling and diving deployment and monitoring logistical accessibility, visual identification of sites from the surface, in combination with tourism return value. Our results suggest that of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region scale), with many unable to maintain positive carbonate budgets under future conditions. Partial dependence analysis revealed a range of depths from 10-15m where the probability of a reef being clustered into the High cover cluster increased substantially. Additionally, two-way partial dependence analysis revealed that high incoming connectivity is important for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth range, and targetting well connected reefs, when assessing possible intervention sites may reduce the exposure of planted and seeded corals to high thermal stress, while still maintaining effective logistical access to sites. Managers are also likely to consider the positive potential for flow on effects due to outgoing connectivity when selecting intervention sites [@hock2017]. While reefs with high outgoing connectivity levels were found to have greater probability of being assigned to low cover clusters when assessed individually (@fig-ts-cluster-pdp-twoway-incoming), within the context of reef depth, reefs with high depths and high outgoing connectivity levels had the greatest probability of being clustered into high cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results suggests that interventions targetting deep reefs will also benefit from high incoming and outgoing connectivity levels, allowing for high potential for site resilience and outgoing flow of benefits and recovery. Transient, and difficult to measure, properties such as reef connectivity are likely to increase uncertainty in intervention planning to a greater degree than environmental characteristics such as reef depth. Overall, reef intervention planning is a complex and challenging decision-making process [@anthony2020], and our results suggest that intervention site depth may be an important factor to additionally consider when assessing the long-term resilience of sites.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is the supply of recruits that a reef receives following disturbance [@hock2017; @mumby2008; @torres2024]. Our results suggest that incoming connectivity contributes positively to high cover cluster allocation for reefs over 10m deep. At depths of less than 10m thermal stress may be too high for juveniles to contribute to long term recovery. While there is interaction between incoming connectivity and depth for reef cluster prediction, according to available input data, there are few reefs in the GBR that occur at depths \>10m and have Log10 weighted incoming connectivity levels below 0 (@sfig-access-esm1-5-mean-carb-budget), resulting in overall low importance of connectivity as an individual feature for clustering across the GBR. Furthermore, our results suggest no direct, first order, relationship between weighted incoming connectivity and the ability of reefs to maintain positive carbonate budgets under future conditions. There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent generation connectivity modelling is able to identify reefs that are important for recovery processes, with findings dependent on the larval connectivity model used. These investigations were performed using prior generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods, without considering the influence of reef depth on resilience. Using updated 1km^2^ resolution larval connectivity across the whole GBR, with a limited temporal sample of particle flows, our results suggest that larval connectivity alone may not be important or reef resilience and recovery, but should be considered in combination with more influential factors such as depth. In-depth comparisons and validation of multiple connectivity model outputs would improve certainty and allow for the targetting of reefs by combining abiotic factors such as depth with connectivity.

Our findings on the importance of connectivity may differ from literature due to the highly uncertain nature of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching events decreasing the breeding capacity of strongly connected reefs. Analysing daily connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean connectivity may not be suitable for capturing the seasonal - ecological time scale connectivity between locations due to high daily variability in connectivity data. This is supported by the connectivity data used in our investigation, which have large relative standard deviations for connectivity between reefs over the 24 days used (@sfig-connectivity-rsd). Using alternative summarisation methods that capture ecological-timescale patterns in coral larval dispersal, such as cumulated implicit connectivity [@vogt-vincent2023], future investigations may find increased importance of both depth and reef connectivity for resilience, in line with other investigations [@hock2017; @torres2024]. Further improvements and research in connectivity modelling are necessary when incorporating connectivity as a supplementary factor in reef resilience and intervention-planning analyses.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced available light levels. [@baker1975; @huston1985; @ruiz-diaz2022]. If included in this investigation it is possible that reef recovery ability, via regrowth, would be reduced at greater depths, reducing the positive effect of depth refugia for temporal resilience. Additionally, reduced growth rates at increased depths may reduce the carbonate budget maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress projections are likely to influence the results obtained in this investigation. While the resolution of connectivity data used in this investigation improves upon data used in comparable studies [@bozec2025], the time-mean implementation used may not effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity implementation methods that better capture the ecological-timescale trends in reef connections rather than the high daily variability, would improve confidence in the connectivity findings of this investigation [@vogt-vincent2023]. Additionally greater year - sample sizes for baseline connectivity data as future computation improvements are made, along with GBR scale validation efforts, will improve confidence in connectivity data for ecological modelling and management applications. The DHW data used in this investigation have not been bias corrected for the Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022] developed using statistical downscaling. Additionally, Lachs et al. [@lachs2021] suggest the utility of DHW levels for prediction of coral bleaching can be improved by reducing the threshold for HotSpot Anomaly values from MMM +1 to MMM and reducing the accumulation window from 12 weeks to 8 weeks. We investigate a range of future climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox to account for many possible futures, and our results suggest the factors that influence reef resilience are consistent across these scenarios.

@@

Improvements in remote sensing approaches for broad scale bathymetry measurement and understanding of how depth impacts experienced thermal stress on reefs will further enhance investigations of depth as a critical factor for reef resilience. Additional analyses could build on the findings of this study by assessing the impacts of prioritising reef depth in reef intervention decision making. Future analyses could incorporate other environmental and ecological stressors such as wave activity, ocean acidification, and Crown-of-Thorn Starfish [@lin2023; @mcclanahan2005; @grimaldi2023; @grimaldi2022; @liu2024; @eyre2018].

@@

The results presented suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of thermal refugia for shallow-water reefs on the GBR. Targetting reefs with median depths greater than 10m and high incoming and ougoing larval connections may aid in identifying the most resilient reefs under climate change. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef intervention decision making, as well as further research and development in connectivity modelling methods for whole-ecosystem scales.
